SMT-Boosted Security Types for Low-Level MPC

### 2024-10-14 esop25/appendix.tex
checkpoint

>>>>

\subsection{Type Checking Soundness}

\begin{lemma}
  \label{lemma-fresh}
  Given $w$ and $\cid$ with $\fresh(w,\cid)$, if $\notg{\eqs}[w/y_1,\cid/y_n]$
  is not satisfiable, then there exists no $w',\cid'$ such that
  $\notg{\eqs}[w'/y_1,\cid'/y_n]$ is satisifiable.
\end{lemma}

\begin{proof}
  Let $\eqs = \notg{\eqs}[w/y_1,\cid/y_n]$ and $\eqs' =
  \notg{\eqs}[w'/y_1,\cid'/y_n]$. If $w',\cid'$ are fresh or chosen
  such that $\eqs'$ is an $\alpha$-renaming of $\eqs$, the result is
  immediate.  Otherwise, it must be the case that at least two
  distinct variables $x_1$ and $x_2$ in $\eqs$ are renamed as a single
  variable $x$ in $\eqs'$, given freshness of $w,\cid$. Let
  $\mathit{rename}$ be the function that maps variables in $\eqs$ to
  variables in $\eqs'$ such that $\mathit{rename}(\eqs) = \eqs'$. Let
  $\store'$ be a model of $\eqs'$, and define $\store$ such that:
  $$
  \forall x \in \dom(\vars(E)) . \store(x) \defeq \store'(\mathit{rename}(x)) 
  $$
  Then $\store$ must be a model of $\eqs$, which is a contradiction. \qed
\end{proof}

\begin{lemma}
  \label{lemma-eqs-notg}
  For all $\fresh(\mv_1,\ldots,\mv_n)$, $\notg{E_1}$, $\notg{E_2}$, assuming: 
  \begin{mathpar}
    \notg{\eqs_1}[\mv_1/y_1]\cdots[\mv_n/y_n]  \redx \eqs_1

    \notg{\eqs_2}[\mv_1/y_1]\cdots[\mv_n/y_n]  \redx \eqs_2

    E_1 \models E_2
  \end{mathpar}
  then there exists no $\mv_1',\ldots,\mv_n'$ such that:
    \begin{mathpar}
    \notg{\eqs_1}[\mv_1'/y_1]\cdots[\mv_n'/y_n]  \redx \eqs_1

    \notg{\eqs_2}[\mv_1'/y_1]\cdots[\mv_n'/y_n]  \redx \eqs_2

    E_1' \not\models E_2'
  \end{mathpar}  
\end{lemma}

\begin{proof}
  Given the assumptions, suppose that $E_1' \not\models E_2'$
  is an $\alpha$-renaming of $E_1 \models E_2$. In this
  case the consequence follows immediately. Otherwise,
  suppose on the contrary we can witness
  $E_1' \not\models E_2'$. By definition this means that
  $E_1' \wedge \neg E_2'$ is satisfiable
  but $E_1 \wedge \neg E_2$ is not, which is a contradiction
  by Lemma \ref{lemma-fresh}. \qed 
\end{proof}

\begin{proof}[Theorem \ref{theorem-mtj}]
  (Sketch) The result follows by induction on derivation of the
  main type judgement for $\instr$. The \TirName{Mesg} case follows
  by Lemma \ref{lemma-atj-sound}. The \TirName{Encode} and \TirName{App}
  cases follow by definition of the \TirName{Sig} rule and Lemma
  \ref{lemma-eqs-notg}. 
\end{proof}

### 2024-10-14 esop25/metalang.tex
checkpoint

The $\metaprot$ language \cite{skalka-near-ppdp24} includes structured
data and function definitions for defining composable protocol
elements at a higher level of abstraction.  The $\metaprot$ language
is a \emph{metalanguage} aka metaprogramming language, where a
$\fedprot$ protocols is the result of computation. In addition to
these declarative benefits of $\metaprot$, component definitions
support compositional verification of larger protocols as we will
discuss with extended examples in Section \ref{section-examples}.

@@

The syntax of $\metaprot$ is defined in Figure
\ref{fig-metaprotsyntax}.  It includes a syntax of values $\mv$ that
include client ids $\cid$, identifier strings $w$, expressions $\be$
in field $\mathbb{F}_p$, record values, and $\minicat$ variables
$x$. $\metaprot$ expression forms allow dynamic construction of these
values. $\metaprot$ \emph{instruction} forms allow dynamic
construction of $\minicat$ protocols $\prog$, incorporating expression
evaluation. The syntax also supports definitions of functions that
compute values $\mv$ and, as a distinct form, functions that compute
protocols.  Formally, we consider a complete metaprogram to include
both a codebase and a ``main'' program that uses the codebase.
We disallow recursion, mainly to guarantee decidability
of type dependence (Section \ref{section-metalangty}).
\begin{definition}
A \emph{codebase} $\codebase$ is a list of function 
declarations. We write $ \codebase(f) = y_1,\ldots,y_n,\ e$
if $f(y_1,\ldots,y_n) \{ e \} \in \codebase$, and we
write  $ \codebase(f) = y_1,\ldots,y_n,\ \instr$
if $f(y_1,\ldots,y_n) \{ \instr \} \in \codebase$.
%A \emph{metaprogram}, aka \emph{metaprotocol}  is a pair of a 
%codebase and expression $\codebase, e$. We may omit
%$\codebase$ if it is clear from context.  
\end{definition}

@@

Our first goal in the $\metaprot$ type theory is to define an
algorithmic system that is sound for both confidentiality and
integrity typings as defined in Sections \ref{section-cpj} and
\ref{section-ipj}. Returning to the example in Section
\ref{section-cpj}, note that the key equivalence of the $\ttt{mux}$
expression with a one-time-pad encryption we observe there
is not trivial. To ensure that this sort of equivalence
can be picked up by the type system, we introduce a
an annotation form that allows the programmer to provide
the needed hint.

Consider the following $\ttt{encode}$ function definition that
generalizes this encoding for any identifier $y$ and sender, receiver
pair $s,r$. The second line in the body is a hint that
the message $\mx{y}{r}$ can be considered equivalent to
$\neg\sx{y}{s} \fplus \rx{y}{s}$:
$$
\begin{array}{l}
\ttt{encode}(y,s,r) \{\\
\quad \xassign{\mx{y}{r}}{\mux{\secret{y}}{\neg\flip{y}}{\flip{y}}}{s};\\
\quad \eqcast{\mx{y}{r}}{\neg\sx{y}{s} \fplus \rx{y}{s}}\\
\}
\end{array}
$$
This hint can be validated using SMT, and then the
syntactic structure of $\neg\sx{y}{s} \fplus \rx{y}{s}$
allows its immediate interpretation as a one-time-pad encoding.

Our second goal in the $\metaprot$ type system is to minimize the
amount of SMT solving needed for type verification.  Returning to the
$\ttt{encode}$ example, type checking in our systems only requires the
hint to be verified once, with the guarantee that $\ttt{encode}$ can
be applied anywhere without needing to re-verify the hint in
application instances. To verify the hint in $\ttt{encode}$, we can
just choose arbitrary values $w,\cid_1,\cid_2$ for $y$, $s$, and $r$,
evaluate $\ttt{encode}(w,\cid_1,\cid_2) \redx \prog$, and then verify:
$$\toeq{\prog} \models \mx{w}{\cid_2} \eop
\neg\sx{w}{\cid_1} \fplus \rx{w}{\cid_1}$$ 
Since $\ttt{encode}$ is closed, validity is guaranteed for
any instance of $y$, $s$, and $r$.

@@

A core element of $\metaprot$ type checking is type checking
of $\minicat$ expressions. The integrity type system presented in Section
\ref{section-ipj} is already algorithmic and ready to use. But confidentiality
typing presented in Section \ref{section-cpj} is not syntax-directed
due to the \TirName{Encode} rule. But as described above, by introducing
hint annotations we can ``cast'' any relevant expression form into
the simplest equivalent syntactic form for one-time-pad encoding.
Thus, in Figure \ref{fig-atj} we present the algorithm for
confidentiality type judgements, where we eliminate the need for
integrated SMT solving by assuming this sort of casting.

@@

Hoare-style types for instructions have the following form:
$$
\hty{\eqs_1}{\Gamma}{R}{\Delta}{\eqs_2}
$$
Here, $\eqs_1$ and $\eqs_2$ are the pre- and post- conditions
respectively, $\Gamma$ is the confidentiality type environment of the
protocol resulting from execution of the instruction, which is sound
wrt confidentiality typing, $R$ are the one-time-pads consumed in the
confidentialty typing, and $\Delta$ is the sound integrity type
environment.

@@

For example, here is a valid typing of $\ttt{encode}$.  It says that
under any precondition, evaluating $\ttt{encode}$ results in a cipher
type for the encoded message $\mx{x}{r}$, which consumes the
one-time-pad $\rx{x}{s}$, and the integrity of $\mx{x}{r}$ is
determined by the integrity of $\rx{x}{s}$ and $\sx{x}{s}$. The
postcondition expresses the key confidentiality property of
$\mx{x}{r}$, but also may be practically useful for correctness
properties since it is a simpler expression form than the $\ttt{mux}$
form:
$$
\ttt{encode} :
\begin{array}[t]{ll}
  \Pi x,s,r . & \{ \} \quad \mx{x}{r} : \cty{\rx{x}{s}}{\sx{x}{s}}, \{ \rx{x}{s} \} \\
  & \phantom{\{ \}} \quad \mx{x}{r} : \ity{s}{\{ \rx{x}{s}, \sx{x}{s} \}} \\
  & \{ \mx{x}{r} \eop \neg\sx{y}{s} \fplus \rx{y}{s} \}
\end{array}
$$
A key property of this example and our type system generally is that once
the $\Pi$ type is verified, typing any application of it requires verification
of the precondition instance, but not the postcondition instance.

@@

We equate types up to evaluation of subexpressions as defined in
Figure \ref{fig-notg}. Since expression evaluation is total we can
just evaluate types of closed instructions, which have closed types,
to obtain Hoare typings for instructions. In Figure \ref{fig-mtj} we
define the type derivation rules for $\metaprot$. In the
\TirName{Mesg} rule we use algorithmic confidentiality and integrity
type checking in a straightforward manner.  In the \TirName{Encode}
rule, we verify the hint given in the type annotation, and then use it
for type checking.  In the \TirName{App} rule we verify preconditions
of the given function type and instantiate the type and postconditions
with the given expression arguments.  The \TirName{Sig} rule we
specify how to verify $\Pi$ types for functions. Note that the values
$\mv_1,\ldots,\mv_n$ chosen to instantiate the function parameters are
chosen ``fresh'', meaning that any strings and client identifiers
occurring as substrings in $\mv_1,\ldots,\mv_n$ are unique and
not used in any program source code. This ensures generality
of verification as demonstrated in Lemma \ref{lemma-eqs-notg}.

@@

\subsection{Type Checking Soundness}

\begin{lemma}
  \label{lemma-fresh}
  Given $w$ and $\cid$ with $\fresh(w,\cid)$, if $\notg{\eqs}[w/y_1,\cid/y_n]$
  is not satisfiable, then there exists no $w',\cid'$ such that
  $\notg{\eqs}[w'/y_1,\cid'/y_n]$ is satisifiable.
\end{lemma}

\begin{proof}
  Let $\eqs = \notg{\eqs}[w/y_1,\cid/y_n]$ and $\eqs' =
  \notg{\eqs}[w'/y_1,\cid'/y_n]$. If $w',\cid'$ are fresh or chosen
  such that $\eqs'$ is an $\alpha$-renaming of $\eqs$, the result is
  immediate.  Otherwise, it must be the case that at least two
  distinct variables $x_1$ and $x_2$ in $\eqs$ are renamed as a single
  variable $x$ in $\eqs'$, given freshness of $w,\cid$. Let
  $\mathit{rename}$ be the function that maps variables in $\eqs$ to
  variables in $\eqs'$ such that $\mathit{rename}(\eqs) = \eqs'$. Let
  $\store'$ be a model of $\eqs'$, and define $\store$ such that:
  $$
  \forall x \in \dom(\vars(E)) . \store(x) \defeq \store'(\mathit{rename}(x)) 
  $$
  Then $\store$ must be a model of $\eqs$, which is a contradiction. \qed
\end{proof}

\begin{lemma}
  \label{lemma-eqs-notg}
  For all $\fresh(\mv_1,\ldots,\mv_n)$, $\notg{E_1}$, $\notg{E_2}$, assuming: 
  \begin{mathpar}
    \notg{\eqs_1}[\mv_1/y_1]\cdots[\mv_n/y_n]  \redx \eqs_1

    \notg{\eqs_2}[\mv_1/y_1]\cdots[\mv_n/y_n]  \redx \eqs_2

    E_1 \models E_2
  \end{mathpar}
  then there exists no $\mv_1',\ldots,\mv_n'$ such that:
    \begin{mathpar}
    \notg{\eqs_1}[\mv_1'/y_1]\cdots[\mv_n'/y_n]  \redx \eqs_1

    \notg{\eqs_2}[\mv_1'/y_1]\cdots[\mv_n'/y_n]  \redx \eqs_2

    E_1' \not\models E_2'
  \end{mathpar}  
\end{lemma}

\begin{proof}
  Given the assumptions, suppose that $E_1' \not\models E_2'$
  is an $\alpha$-renaming of $E_1 \models E_2$. In this
  case the consequence follows immediately. Otherwise,
  suppose on the contrary we can witness
  $E_1' \not\models E_2'$. By definition this means that
  $E_1' \wedge \neg E_2'$ is satisfiable
  but $E_1 \wedge \neg E_2$ is not, which is a contradiction
  by Lemma \ref{lemma-fresh}. \qed 
\end{proof}

The following Theorem establishes our main result, which
is the $\metaprot$ type checking is sound with respect to both
confidentiality and integrity typings. Since these typings
imply security hyperproperties of confidentiality and integrity,
$\metaprot$ type checking implicitly enforces them.
\begin{theorem}[Soundness of $\metaprot$ type checking]
  \label{theorem-mtj}
  Given preprocessing predicate $\eqspre$, program $\instr$, and verified $\tsig$, if
  the judgement: $$\mtj{\instr}{\eqspre}{\Gamma}{R}{\Delta}{\eqs}$$ is derivable then
  $\instr \redx \prog$ and:
  \begin{enumerate}
  \item $\cpj{R}{\eqs}{\eqspre \wedge \toeq{\prog}}{\Gamma}$ is valid.
  \item $\ipj{\eqs}{\prog}{\Delta}$ is valid.
  \end{enumerate}
\end{theorem}

\begin{proof}
  (Sketch) The result follows by induction on derivation of the
  main type judgement for $\instr$. The \TirName{Mesg} case follows
  by Lemma \ref{lemma-atj-sound}. The \TirName{Encode} and \TirName{App}
  cases follow by definition of the \TirName{Sig} rule and Lemma
  \ref{lemma-eqs-notg}. 
\end{proof}

>>>>

The $\metaprot$ language \cite{skalka-near-ppdp24} includes structured
data and function definitions for defining composable protocol
elements at a higher level of abstraction than $\minicat$.  The
$\metaprot$ language is a \emph{metalanguage} aka metaprogramming
language, where a $\fedprot$ protocols is the result of
computation. In addition to these declarative benefits of $\metaprot$,
component definitions support compositional verification of larger
protocols. Separate verification of well-designed components results
in confidentiality and integrity properties to be recorded in their
types, allowing for significant reduction of SMT verification in whole
program analysis, as we will discuss with extended examples in Section
\ref{section-examples}.

@@

The syntax of $\metaprot$ is defined in Figure
\ref{fig-metaprotsyntax}.  It includes a syntax of values $\mv$ that
include client ids $\cid$, identifier strings $w$, expressions $\be$
in field $\mathbb{F}_p$, record values, and $\minicat$ variables
$x$. $\metaprot$ expression forms allow dynamic construction of these
values. $\metaprot$ \emph{instruction} forms allow dynamic
construction of $\minicat$ protocols $\prog$, incorporating expression
evaluation. The syntax also supports definitions of functions that
compute values $\mv$ and, as a distinct form, functions that compute
protocols $\prog$.  Formally, we consider a complete metaprogram to include
both a codebase and a ``main'' program that uses the codebase.
We disallow recursion, mainly to guarantee decidability
of type dependence (Section \ref{section-metalangty}).
\begin{definition}
A \emph{codebase} $\codebase$ is a list of function 
declarations. We write $ \codebase(f) = y_1,\ldots,y_n,\ e$
if $f(y_1,\ldots,y_n) \{ e \} \in \codebase$, and we
write  $ \codebase(f) = y_1,\ldots,y_n,\ \instr$
if $f(y_1,\ldots,y_n) \{ \instr \} \in \codebase$.
%A \emph{metaprogram}, aka \emph{metaprotocol}  is a pair of a 
%codebase and expression $\codebase, e$. We may omit
%$\codebase$ if it is clear from context.  
\end{definition}

@@

Our first goal in the $\metaprot$ type theory is to define an
algorithmic system that is sound for both confidentiality and
integrity typings as defined in Sections \ref{section-cpj} and
\ref{section-ipj}. Returning to the YGC secret encoding example in
Section \ref{section-cpj}, note that the key equivalence of the
$\ttt{mux}$ expression with a one-time-pad encryption we observe there
is not trivial. To ensure that this sort of equivalence can be picked
up by the type system, we introduce an $\ttt{as}$ annotation form that
allows the programmer to provide the needed hint.

Consider the following $\ttt{encode}$ function definition that
generalizes this YGC encoding for any identifier $y$ and sender, receiver
pair $s,r$. The second line in the body is a hint that
the message $\mx{y}{r}$ can be considered equivalent to
$\neg\sx{y}{s} \fplus \rx{y}{s}$:
$$
\begin{array}{l}
\ttt{encode}(y,s,r) \{\\
\quad \xassign{\mx{y}{r}}{\mux{\secret{y}}{\neg\flip{y}}{\flip{y}}}{s};\\
\quad \eqcast{\mx{y}{r}}{\neg\sx{y}{s} \fplus \rx{y}{s}}\\
\}
\end{array}
$$
This hint can be validated using SMT, and then the
syntactic structure of $\neg\sx{y}{s} \fplus \rx{y}{s}$
allows its immediate interpretation as a one-time-pad encoding.

Our second goal in the $\metaprot$ type system is to minimize the
amount of SMT solving needed for type verification.  Returning to the
$\ttt{encode}$ example, type checking in our systems only requires the
hint to be verified once, with the guarantee that $\ttt{encode}$ can
be applied anywhere without needing to re-verify the hint in
application instances. To verify the hint in $\ttt{encode}$, we can
just choose arbitrary ``fresh'' values $w,\cid_1,\cid_2$ for $y$, $s$,
and $r$, evaluate $\ttt{encode}(w,\cid_1,\cid_2) \redx \prog$, and
then verify:
$$\toeq{\prog} \models \mx{w}{\cid_2} \eop
\neg\sx{w}{\cid_1} \fplus \rx{w}{\cid_1}$$ 
Since $\ttt{encode}$ is closed, validity is guaranteed for
any instance of $y$, $s$, and $r$.

@@

A core element of $\metaprot$ type checking is type checking
of $\minicat$ expressions. The integrity type system presented in Section
\ref{section-ipj} is already algorithmic and ready to use. But confidentiality
typing presented in Section \ref{section-cpj} is not syntax-directed
due to the \TirName{Encode} rule. But as described above, by introducing
hint annotations we can ``coerce'' any relevant expression form into
the simplest equivalent syntactic form for one-time-pad encoding.
Thus, in Figure \ref{fig-atj} we present the algorithm for
confidentiality type judgements, where we eliminate the need for
integrated SMT solving by assuming this sort of casting.

@@

Hoare-style types for instructions have the following form:
$$
\hty{\eqs_1}{\Gamma}{R}{\Delta}{\eqs_2}
$$
Here, $\eqs_1$ and $\eqs_2$ are the pre- and postconditions
respectively, $\Gamma$ is the confidentiality type environment of the
protocol resulting from execution of the instruction, which is sound
wrt confidentiality typing, $R$ are the one-time-pads consumed in the
confidentialty typing, and $\Delta$ is the sound integrity type
environment.

@@

For example, here is a valid typing of $\ttt{encode}$.  It says that
under any precondition, evaluating $\ttt{encode}$ results in a cipher
type for the encoded message $\mx{x}{r}$, which consumes the
one-time-pad $\rx{x}{s}$, and the integrity of $\mx{x}{r}$ is
determined by the security level of $s$. The
postcondition expresses the key confidentiality property of
$\mx{x}{r}$, but also may be practically useful for correctness
properties since it is a simpler expression form than the $\ttt{mux}$
form:
$$
\ttt{encode} :
\begin{array}[t]{ll}
  \Pi x,s,r . & \{ \}\quad \mx{x}{r} : \cty{\rx{x}{s}}{\sx{x}{s}}, \{ \rx{x}{s} \} \ \cdot\\
  & \phantom{\{ \}} \quad \mx{x}{r} : \ity{s}{\setit{}} \\
  & \{ \mx{x}{r} \eop \neg\sx{y}{s} \fplus \rx{y}{s} \}
\end{array}
$$
A key property of this example and our type system generally is that once
the $\Pi$ type is verified, typing any application of it requires verification
of the precondition instance, but not the postcondition instance.

@@

We equate types up to evaluation of subexpressions as defined in
Figure \ref{fig-notg}. Since expression evaluation is total we can
just evaluate types of closed instructions, which are guaranteed to
have closed types, to obtain Hoare typings for instructions. In Figure
\ref{fig-mtj} we define the type derivation rules for $\metaprot$. In
the \TirName{Mesg} rule we use algorithmic confidentiality and
integrity type checking in a straightforward manner.  In the
\TirName{Encode} rule, we verify the hint given in the type
annotation, and then use it for type checking.  In the \TirName{App}
rule we verify preconditions of the given function type and
instantiate the type and postconditions with the given expression
arguments.  The \TirName{Sig} rule we specify how to verify $\Pi$
types for functions. Note that the values $\mv_1,\ldots,\mv_n$ chosen
to instantiate the function parameters are chosen ``fresh'', meaning
that any strings and client identifiers occurring as substrings in
$\mv_1,\ldots,\mv_n$ are unique and not used in any program source
code. This ensures generality of verification as demonstrated in Lemma
\ref{lemma-eqs-notg} in the Appendix.

@@

The following Theorem establishes our main result, that the
$\metaprot$ type checking is sound with respect to both
confidentiality and integrity typings. Since these typings imply
security hyperproperties of confidentiality and integrity, $\metaprot$
type checking implicitly enforces them. Proof details are given in the
Appendix.
\begin{theorem}[Soundness of $\metaprot$ type checking]
  \label{theorem-mtj}
  Given preprocessing predicate $\eqspre$, program $\instr$, and verified $\tsig$, if
  the judgement: $$\mtj{\instr}{\eqspre}{\Gamma}{R}{\Delta}{\eqs}$$ is derivable then
  $\instr \redx \prog$ and:
  \begin{enumerate}
  \item $\cpj{R}{\eqs}{\eqspre \wedge \toeq{\prog}}{\Gamma}$ is valid.
  \item $\ipj{\eqs}{\prog}{\Delta}$ is valid.
  \end{enumerate}
\end{theorem}
